---
title : "Insert Title Here"
author: "Morde Bernstein, Marley McGehee, Kimberly McNeil"
date: "5/01/2026"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      message = TRUE)
library(tidyverse)
library(knitr)

```


## Introduction

Our first dataset is focused on adult asthma prevalence by state in 2020. We obtained this dataset from the CDC website at the [2020 Archived State or Territory Asthma Data](https://www.cdc.gov/asthma/archivedata/2020/2020_archived_states_territory.html) page. The data has a sample size of 50 and includes both numbers with current asthma and percent with current asthma variables. 

Our second dataset is about air pollution particulate matter by state, also in 2020. We obtained this data from the National Institute on Minority Health and Health Disparities at the [Physical Environment - Air Pollution] (https://hdpulse.nimhd.nih.gov/data-portal/physical/table?age=001&age_options=ageall_1&demo=234&demo_options=air_pollution_1&physicaltopic=002&physicaltopic_options=physical_2&race=00&race_options=raceall_1&sex=0&sex_options=sexboth_1&statefips=00&statefips_options=area_states) page. The sample size is 50, and the data includes air particulates in micrograms per cubic meter (PM2.5). This is a measure of air pollution which looks at the average number of particulates in the air daily.

## Data Import

```{r}
# Dataset 1 Import
asthma <- read_csv("asthma_prevalence - Sheet1.csv")

# Initialize cleaning the column names (and giving names to the unnamed)
asthma <- janitor::clean_names(asthma) 

# Make the janitor's work clearer
asthma <- asthma %>%
  rename(
    state = state_or_territory,
    asthma_number = number_with_current_asthma,
    asthma_percent = percent_with_current_asthma_se)

# Dataset 2 Import

pollution <- read_csv("HDPulse_data_export (1).csv", 
    skip = 5)

# Initialize cleaning the column names (and giving names to the unnamed)
pollution <- janitor::clean_names(pollution) 

# Make the janitor's work clearer
pollution <- pollution %>%
  rename(
    state = state,
    fips = fips,
    pollution = micrograms_per_cubic_meter_pm2_5_1)

```

```{r}
asthma_air <- asthma %>% left_join(pollution, join_by(state)) %>%
  relocate(fips, .after = state)
```

## Exploratory Data Analysis

## Research

## Conclusion


